Understanding the Impact of Socioeconomic and Regional Factors on Health and Educational Outcomes

### 2024-12-15 paper.qmd
final with edits

This study examines the relationships between socioeconomic, health, and education indicators in U.S. counties using data from the American Community Survey, the USDA Food Environment Atlas, Small Area Income and Poverty Estimates, and other sources.We analyze how factors such as residential segregation, health insurance coverage, access to healthy foods, and region influence life expectancy, and how residential segregation and region affect school funding adequacy. Our findings reveal that life expectancy is influenced by residential segregation and uninsured rates, though these effects are minor compared to the more substantial influence of household income and regional differences. Life expectancy averages are highest in the Northeast, followed by the Midwest, West, and South. Region also plays the most significant role out of all included variables in determining the likelihood of school funding adequacy, with schools in the Northeast and Midwest being far more likely to receive adequate funding than those in the South. Other factors, such as residential segregation, property taxes, and household income, show statistically significant but smaller effects. These results highlight the complex interplay between socioeconomic variables and geographic location in shaping public well-being. By identifying key disparities in health and education, this study underscores the importance of equitable resource allocation and targeted policies to address regional and socioeconomic inequalities.

@@

Life expectancy and school funding are important indicators of public well-being because they capture fundamental aspects of health, education, and social equity that are essential for the development of individuals and communities. Life expectancy is an important health indicator that reflects not only the overall health of a society, but also its stability, economic resilience, and levels of inequality. While access to quality healthcare and nutritious foods are drivers of well-being, where you live and the nature of the community you live in likely also have an influence on health outcomes such as life expectancy. According to "The Root Causes of Health Inequity - Communities in Action" report from the National Institutes of Health, we can infer that differences in life expectancy between regions and socioeconomic groups often reveal systemic inequalities, with marginalized communities experiencing shorter life spans due to limited access to health care, healthy living environments, and economic opportunities.^1^ When it comes to educational indicators, adequately funded schools ensure that students have equal access to learning opportunities, serving as a crucial investment in social mobility. The Stanford Center for Opportunity Policy in Education highlights that "the unequal allocation and inadequate levels of resources in schools and communities \[are\] at the heart of many gaps in student opportunity."^2^ This emphasizes the importance of equitable and adequate school funding in closing opportunity gaps and empowering students from disadvantaged backgrounds to achieve academic and economic success. By better understanding the correlates of school funding adequacy, we can take a major step towards implementing policy that targets the areas and communities most affected by this issue.

@@

Variables of interest were selected a priori based on their perceived relevance to the two outcomes we wanted to better understand: average life expectancy and school funding adequacy. While nutrition and access to quality healthcare are inherently linked to general wellbeing, it is worthwhile to assess how influential the environmental and financial aspects of these factors are on our tangible outcome. Therefore, we selected two key predictors for our life expectancy model: the percentage of the population with limited access to healthy foods and the percentage of uninsured adults in a county. The former is defined by the US Department of Agriculture as "the percentage of the \[county\] population that is low-income and does not live close to a grocery store", where "close" is defined differently for rural and nonrural areas.^6^ The latter variable represents the percentage of adults in a county younger than 65 who lack health insurance. We also were interested in seeing how residential segregation would impact life expectancy among these predictors, given existing studies on the negative health effects of segregation, and whether this relationship would exhibit any variations across different regions. This is why we chose to include residential segregation index values as a predictor. The residential segregation index, used by the County Health Rankings & Roadmaps program, measures how evenly two groups, such as Black and white residents, are distributed across geographic areas, with values ranging from 0 (complete integration) to 100 (complete segregation).^7^

@@

### *Data Overview*

@@

Based on the multiple linear regression results, residential segregation, percentage of uninsured adults, region, and median household income all showed statistically significant effects on life expectancy. Specifically, for every \$10,000 increase in median household income, life expectancy increases by 1.25 years on average. In addition, the impact of region on life expectancy was substantial: life expectancy is on average 1.96 years higher in the West region compared to the South, 2.92 years higher in the Northeast compared to the South, and 1.92 years higher in the Midwest compared to the South, even when controlling for income, uninsured population, and racial diversity. This suggests, income as a factor withstanding, the differing quality of and access to healthcare and public services found in each region can have a significant impact on life expectancy.

Surprisingly, holding all other variables constant, for every 1 unit increase in percentage of uninsured adults in a county, a county's average life expectancy increases by 0.032 years. While statistically significant, many studies covering the relationship between an individual’s health and having insurance coverage have proven its positive correlation, suggesting our results may benefit from a closer look. In assessing our dataset, as seen in Appendix 1A, we find that our dataset contains more observations for counties with low uninsured individuals, or equivalently, counties where most people are covered. As the timeframe of our dataset is 2018-2022, it is likely that the aforementioned observations require greater scrutiny of the significance of the impact of this variable on average life expectancy of a county.

The percentage of population with limited access to healthy foods predictor was not statistically significant in our model, suggesting that the effect of this environmental factor is minor or intertwined with economic variables, such as income, rather than being a direct or standalone influence. In terms of residential segregation, for every 1 unit increase in a county's index value (corresponding to increased segregation), its average life expectancy increases by 0.0071 years in the South, increases by 0.0068 years in the West, decreases by 0.0055 years in the Northeast, and decreases by 0.0004 years in the Midwest, as seen in Figure 1. However, these regional differences are not statistically significant at the 0.05 significance level given the high p-values of all residential segregation-region interaction terms. We confirmed this result with a nested F-test, comparing a full model to a reduced model that excluded the interaction term between region and residential segregation. This did not show a statistically significant difference in RSS (p=0.624), indicating that the interaction between region and residential segregation does not add meaningful explanatory power to our model, Consequently, we did not find sufficient evidence to support regional variations in the relationship between segregation and life expectancy. Instead, based on residential segregation's coefficient estimate in the model, life expectancy does appear to increase overall with an increase in residential segregation, however minor this increase may be. One possible theory for this observation is that counties with more racial uniformity might experience less racial conflict and more social cohesion, especially for minority communities.

@@

The adjusted R-squared value for the model was 0.6065, indicating that a majority of the variability in life expectancy was explained by the socioeconomic and racial predictors included in the model. In conducting diagnostics on the multilinear regression model, we first checked to see if the model met assumptions of linearity, independence, normality, and homoscedasticity. The residuals vs. fitted plot showed relatively equal variance and no discernible pattern, meaning that the linearity and homoscedasticity assumptions were reasonably met. In addition, because each observation in the dataset represented a unique county in the U.S., independence could be reasonably assumed. Assessing the Q-Q Plot for normality, the residual distribution was approximately normal, despite a slight divergence on both tails of the plot. Cook's distance was also assessed using the residuals vs. leverage plot; no points were found to be influential. There was a high leverage point (Mineral County, Nevada), and refitting the model without this point yielded no substantial changes in the coefficient estimates for statistically significant predictors; however, the p-value for residential segregation increased from 0.031 to 0.054. Finally, with a VIF \<2 for all predictors, there was little to no multicollinearity among the predictors.

@@

Table 3 below displays the odds ratio results from the logistic regression model for school funding adequacy, with the South as the baseline category for the categorical region variable. While all estimates are statistically significant at the 0.05 significance level, exponentiated regression coefficients from the logistic model demonstrate median household income and median property tax of the county have almost no impact on the odds of a school being adequately funded. Our results demonstrate that a one unit increase to the residential segregation index increases the odds of school funding adequacy by 0.02 times, allowing us to conclude that residential segregation has a minimal effect on the odds of the school being adequately funded. Figure 2 demonstrates an interesting trend–each region shows a different relationship between the residential segregation distribution of adequately and inadequately funded schools, pointing to a potential interaction between region and segregation.

Region, however, has a substantial impact on the odds of school funding adequacy.  Being located in the Northeast or the Midwest has the greatest impact on how well the school is funded, with the odds of adequate school funding being over 200 times higher in a Northeast county compared to a Southern county, and the odds of adequate school funding being over 3 times higher in a Northeast county compared to a Southern county, holding all other variables constant. Such disparities across the country are caused by a number of factors, including: (1) capacity - how well off a state is based on its economy and resources, and (2) effort - the states willingness to provide funding for education. Wealthier states with a high fiscal capacity, (typically those in the Northeast), have more funding available to spend on education than states with more limited resources (typically those in the South and the West).

@@

To assess logistic regression assumptions, we first plotted each numerical predictor against the log odds of school funding adequacy. All plots showed a mostly linear relationship, indicating that the linearity assumption was largely met; interestingly, different linear trends were observed within each plot based on region, pointing to potential interaction effects between region and the three numerical predictors. GVIF for all predictors ranged between 1 and 3, suggesting that there could be moderate multicollinearity between predictors, but not enough to warrant a reconsideration of included variables. Cook's distance was also less than 0.05 for all points, meaning that there were no highly influential observations in the model. Evaluating our model, we found that the logistic regression model demonstrated a strong overall performance with an accuracy rate of 80.33%. The model had a high sensitivity of 91%, indicating that it accurately predicted 91% of actual positive cases of adequate school funding when evaluated on our dataset. However, it had a 60% specificity rate, highlighting it correctly identified cases of inadequate school funding only 60% of the time. Overall, looking at the ROC curve below, the AUC is 0.858, indicating a moderately strong discriminate ability.

@@

In our analysis, we observed the significant impact of the regional location of a county on both its average life expectancy (a health indicator) and school funding adequacy (an educational indicator)--two key components of societal and an individual’s well-being. Life expectancy is highest in the Northeast, followed by the Midwest, West, and South, while school funding adequacy is significantly greater in the Northeast and Midwest compared to the South. Our findings highlight substantial differences in both outcomes between regions, which are often caused by different economic capabilities and different allocations of resources among countries. This study provides empirical evidence that underscores the importance for equitable social policies and targeted investments to address regional disparities and promote well-being across all communities.

@@

Future studies could incorporate additional social variables to enhance the analysis and explore causal relationships between region and outcomes. For example, integrating a dataset on employment rates and school performance could offer deeper insights into how socioeconomic factors shape life expectancy and educational funding, complementing existing variables in our models. Another direction for future work is to delve deeper into understanding the surprising findings regarding the effects of uninsured adult percentage and residential segregation on life expectancy. Expanding the dataset to include a temporal scope, and finding more complete sources of data on variables of interest so that no states have to be excluded from the analysis, would also contribute to a more comprehensive understanding.

>>>>

This study examines the relationships between socioeconomic, health, and educational indicators in U.S. counties using data from the County Health Rankings & Roadmaps program and IPUMS. We analyze how factors such as residential segregation, health insurance coverage, access to healthy foods, and region impact life expectancy, and how residential segregation and region affect school funding adequacy. Our findings reveal that life expectancy is influenced by residential segregation and uninsured rates, though these effects are minor compared to the more substantial influence of household income and regional differences. Life expectancy averages are highest in the Northeast, followed by the Midwest, West, and South. Region also plays the most significant role out of all included variables in determining the likelihood of school funding adequacy, with schools in the Northeast and Midwest being far more likely to receive adequate funding than those in the South. Other factors, such as residential segregation, property taxes, and household income, show statistically significant but smaller effects. These results highlight the complex interplay between socioeconomic variables and geographic location in shaping public well-being. By identifying key disparities in health and education, this study underscores the importance of equitable resource allocation and targeted policies to address regional and socioeconomic inequalities.

@@

Life expectancy and school funding are important indicators of public well-being because they capture fundamental aspects of health, education, and social equity that are essential for the development of individuals and communities. Life expectancy is an important indicator that reflects not only the overall health of a society, but also its stability, economic resilience, and levels of inequality. While access to quality healthcare and nutritious foods are drivers of well-being, where you live and the nature of the community you live in likely also have an influence on health outcomes such as life expectancy. According to "The Root Causes of Health Inequity - Communities in Action" report from the National Institutes of Health, we can infer that differences in life expectancy between regions and socioeconomic groups often reveal systemic inequalities, with marginalized communities experiencing shorter life spans due to limited access to health care, healthy living environments, and economic opportunities.^1^ When it comes to educational indicators, adequately funded schools ensure that students have equal access to learning opportunities, serving as a crucial investment in social mobility. The Stanford Center for Opportunity Policy in Education highlights that "the unequal allocation and inadequate levels of resources in schools and communities \[are\] at the heart of many gaps in student opportunity."^2^ This emphasizes the importance of equitable and adequate school funding in closing opportunity gaps and empowering students from disadvantaged backgrounds to achieve academic and economic success. By better understanding the correlates of school funding adequacy, we can take a major step towards implementing policy that targets the areas and communities most affected by this issue.

@@

Variables of interest were selected *a priori* based on their perceived relevance to the two outcomes we wanted to better understand: average life expectancy and school funding adequacy. While nutrition and access to quality healthcare are inherently linked to general wellbeing, it is worthwhile to assess how influential the environmental and financial aspects of these factors are on our tangible outcome. Therefore, we selected two key predictors for our life expectancy model: the percentage of the population with limited access to healthy foods and the percentage of uninsured adults in a county. The former is defined by the US Department of Agriculture as "the percentage of the \[county\] population that is low-income and does not live close to a grocery store", where "close" is defined differently for rural and nonrural areas.^6^ The latter variable represents the percentage of adults in a county younger than 65 who lack health insurance. We also were interested in seeing how residential segregation would impact life expectancy among these predictors, given existing studies on the negative health effects of segregation, and whether this relationship would exhibit any variations across different regions. This is why we chose to include residential segregation index values as a predictor. The residential segregation index, used by the County Health Rankings & Roadmaps program, measures how evenly two groups, such as Black and white residents, are distributed across geographic areas, with values ranging from 0 (complete integration) to 100 (complete segregation).^7^

@@

### Data Overview

@@

Based on the multiple linear regression results, residential segregation, percentage of uninsured adults, region, and median household income all showed statistically significant effects on life expectancy. Specifically, for every \$10,000 increase in median household income, life expectancy increases by 1.25 years on average, holding other variables constant. In addition, the impact of region on life expectancy was substantial: life expectancy is on average 1.96 years higher in the West region compared to the South, 2.92 years higher in the Northeast compared to the South, and 1.92 years higher in the Midwest compared to the South, even when controlling for income, uninsured population, and racial diversity. This suggests, income as a factor withstanding, the differing quality of and access to healthcare and public services found in each region can have a significant impact on life expectancy.

Surprisingly, holding all other variables constant, for every 1 unit increase in percentage of uninsured adults in a county, a county's average life expectancy increases by 0.032 years. While this was statistically significant, many studies covering the relationship between an individual’s health and having insurance coverage have proven its positive correlation, suggesting our results may benefit from a closer look. In assessing our dataset, as seen in Appendix A1, we find that our dataset contains more observations for counties with low uninsured individuals, or equivalently, counties where most people are covered. As the timeframe of our dataset is 2018-2022, it is likely that the aforementioned observations require greater scrutiny of the significance of the impact of this variable on average life expectancy of a county.

The percentage of population with limited access to healthy foods predictor was not statistically significant in our model, suggesting that the effect of this environmental factor is minor or intertwined with economic variables, such as income, rather than being a standalone influence. In terms of residential segregation, for every 1 unit increase in a county's index value (corresponding to increased segregation), its average life expectancy increases by 0.0071 years in the South, increases by 0.0068 years in the West, decreases by 0.0055 years in the Northeast, and decreases by 0.0004 years in the Midwest, as seen in Figure 1. However, these regional differences are not statistically significant at the 0.05 significance level given the high p-values of all residential segregation-region interaction terms. We confirmed this result with a nested F-test, comparing a full model to a reduced model that excluded the interaction term between region and residential segregation. This did not show a statistically significant difference in RSS (*p*=0.624), indicating that the interaction between region and residential segregation does not add meaningful explanatory power to our model, Consequently, we did not find sufficient evidence to support regional variations in the relationship between segregation and life expectancy. Instead, based on residential segregation's coefficient estimate in the model, life expectancy does appear to increase overall with an increase in residential segregation, however minor this increase may be. One possible theory for this observation is that counties with more racial uniformity might experience less racial conflict and more social cohesion, especially for minority communities, but this should be further investigated in future studies.

@@

The adjusted R-squared value for the model was 0.6065, indicating that a majority of the variability in life expectancy was explained by the socioeconomic and racial predictors included in the model. In conducting diagnostics on the multilinear regression model, we first checked to see if the model met assumptions of linearity, independence, normality, and homoscedasticity. The residuals vs. fitted plot showed relatively equal variance and no discernible pattern, meaning that the linearity and homoscedasticity assumptions were reasonably met. In addition, because each observation in the dataset represented a unique county in the U.S., independence could be reasonably assumed. Assessing the Q-Q Plot for normality, the residual distribution was approximately normal, despite a slight divergence on both tails of the plot. Cook's distance was also assessed using the residuals vs. leverage plot; no points were found to be influential. There was a high leverage point (Mineral County, Nevada), and refitting the model without this point yielded no substantial changes in the coefficient estimates for statistically significant predictors; however, the *p*-value for residential segregation increased from 0.031 to 0.054. Finally, with a VIF \<2 for all predictors, there was little to no multicollinearity among the predictors.

@@

Table 3 below displays the odds ratio results from the logistic regression model for school funding adequacy, with the South as the baseline category for the categorical region variable. While all estimates are statistically significant at the 0.05 significance level, exponentiated regression coefficients from the logistic model demonstrate that a county's median household income and median property tax have almost no impact on the odds of its schools being adequately funded. Our results demonstrate that a one unit increase to the residential segregation index increases the odds of school funding adequacy by 0.02 times, allowing us to conclude that residential segregation has a minimal effect on the odds of the school being adequately funded. Figure 2 demonstrates an interesting trend—each region shows a different relationship between the residential segregation distribution of adequately and inadequately funded schools, pointing to a potential interaction between region and segregation.

Region has a substantial impact on the odds of school funding adequacy. Being located in the Northeast or the Midwest has the greatest impact on how well a school is funded, with the odds of adequate school funding being over 200 times higher in a Northeast county compared to a Southern county, and the odds of adequate school funding being over 3 times higher in a Midwest county compared to a Southern county, holding all other variables constant. Such disparities across the country are caused by a number of factors, including: (1) capacity - how well off a state is based on its economy and resources, and (2) effort - the state's willingness to provide funding for education. Wealthier states with a high fiscal capacity, (typically those in the Northeast), have more funding available to spend on education than states with more limited resources (typically those in the South and the West).

@@

To assess logistic regression assumptions, we first plotted each numerical predictor against the log odds of school funding adequacy. All plots showed a mostly linear relationship, indicating that the linearity assumption was largely met; interestingly, different linear trends were observed within each plot based on region, further supporting potential interaction effects between region and the three numerical predictors. GVIF for all predictors ranged between 1 and 3, suggesting that there could be moderate multicollinearity between predictors, but not enough to warrant a reconsideration of included variables. Cook's distance was also less than 0.05 for all points, meaning that there were no highly influential observations in the model. Evaluating our model, we found that the logistic regression model demonstrated a strong overall performance with an accuracy rate of 80.33%. The model had a high sensitivity of 91%, indicating that it accurately predicted 91% of actual positive cases of adequate school funding when evaluated on our dataset. However, it had a 60% specificity rate, highlighting it correctly identified cases of inadequate school funding only 60% of the time. Overall, looking at the ROC curve below, the AUC is 0.858, indicating a moderately strong discriminate ability.

@@

In our analysis, we observed the significant impact of the regional location of a county on both its average life expectancy (a health indicator) and school funding adequacy (an educational indicator)--two key components of societal and individual well-being. Life expectancy is highest in the Northeast, followed by the Midwest, West, and South, while school funding adequacy is significantly greater in the Northeast and Midwest compared to the South. Our findings highlight substantial differences in both outcomes between regions, which are often caused by different economic capabilities and different allocations of resources among countries. This study provides empirical evidence that underscores the importance for equitable social policies and targeted investments to address regional disparities and promote well-being across all communities.

@@

Future studies could look at additional interaction effects with region and incorporate more social variables to enhance the analysis. For example, integrating a dataset on employment rates and school performance could offer deeper insights into how socioeconomic factors shape life expectancy and educational funding, complementing existing variables in our models. Another direction for future work is to delve deeper into understanding the surprising findings regarding the effects of uninsured adult percentage and residential segregation on life expectancy. Expanding the dataset to include a temporal scope, and finding more complete sources of data on variables of interest so that no states have to be excluded from the analysis, would also contribute to a more comprehensive understanding.
